Adaptive Exponentiated Gradient

### 2014-01-27 lp/ftrl-aux.tex
mostly fixed equation formatting
proofread a first draft, now just need to add a conclusion

\section{Choosing $m_t$ via Follow the Regularized Leader} 
\label{sec:ftrl-aux}
In the previous section we introduced an adaptive exponentiated 
gradient algorithm that achieved a regret bound of 
\[ \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_{t,i})^2 \]
for any sequence $m_t$. We now turn our attention to the task of 
optimizing $m_t$. Clearly, the optimal value of $m_t$ is $m_t = z_t$. However, 
this would require perfect knowledge of the future. Suppose we are less ambitious 
and only want to be competitive with the best fixed value of $m_t$. This optimal value 
is $m^* \eqdef \frac{1}{T} \sum_{t=1}^T z_t$, but we still have the issue that 
we do not $m^*$ ahead of time. However, we have now cast the problem in the familiar 
regret minimzation framework, and can use the machinery of follow the regularized 
leader \cite{hazan2011ftrl}:
\begin{proposition}
\label{prop:ftrl-aux}
Suppose that we choose $m_{t,i} = \frac{1}{t} \sum_{s=1}^{t-1} z_{s,i}$. 
Then, for all $i$, 
\[ \sum_{t=1}^T (z_{t,i} - m_{t,i})^2 \leq \sum_{t=1}^T (z_{t,i} - m^*_i)^2 + 4\log(T+1). \]
\end{proposition}
\begin{proof}
TODO
\end{proof}
Combining Proposition~\ref{prop:ftrl-aux} with Proposition~\ref{prop:aeg} yields 
the algorithm titled EG-Variance in Table~\ref{tab:algorithms}. This algorithm has the 
same updates as an algorithm given by \cite{hazan2010variation}, but uses the 
optimistic prediction $\exp(\theta_t - \eta m_t)$ rather than $\exp(\theta_t)$. 
As a straightforward consequence of Propositions~\ref{prop:aeg} and~\ref{prop:ftrl-aux}, we 
obtain the following regret bound, which improves the bound presented in 
\cite{hazan2010variation}:
\begin{corollary}
\label{cor:EG-variance}
The EG-Variance algorithm achieves a regret bound of
%\[ \Regret(w) \leq \frac{1 + (\log(n)-1)\|w\|_1 + \sum_{i=1}^n w_i\log(w_i)}{\eta} + \eta \left[ \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_i^*)^2 + 4\|w\|_1\log(T+1) \right]. \]
\[ \Regret(w) \leq \frac{\log(n)}{\eta} + \eta \left[ \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_i^*)^2 + 4\log(T+1) \right]. \]
\end{corollary}

>>>>

\section{Choosing $m_t$ via Follow the Regularized Leader} 
\label{sec:ftrl-aux}
In the previous section we introduced an adaptive exponentiated 
gradient algorithm that achieved a regret bound of 
\[ \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_{t,i})^2 \]
for any sequence $m_t$. We now turn our attention to the task of 
optimizing $m_t$. Clearly, the optimal value of $m_t$ is $m_t = z_t$. However, 
this would require perfect knowledge of the future. Suppose we are less ambitious 
and only want to be competitive with the best fixed value of $m_t$. This optimal value 
is $m^* \eqdef \frac{1}{T} \sum_{t=1}^T z_t$, but we still have the issue that 
we do not know $m^*$ ahead of time. However, we have now cast the problem in the familiar 
regret minimzation framework, and can use the machinery of follow the regularized 
leader \cite{hazan2011ftrl} to prove:
\begin{proposition}
\label{prop:ftrl-aux}
Suppose that we choose $m_{t,i} = \frac{1}{t} \sum_{s=1}^{t-1} z_{s,i}$ and that 
$\|z_s\|_{\infty} \leq 1$. Then for all $i$ we have
\[ \sum_{t=1}^T (z_{t,i} - m_{t,i})^2 \leq 2\sum_{t=1}^T (z_{t,i} - m^*_i)^2 + 6. \]
\end{proposition}
\begin{proof}
Note that $m_{t,i}$ is the minimizer of $m_i^2 + \sum_{s=1}^{t-1} (z_{s,i} - m_i)^2$. 
Therefore, by the FTRL lemma, we have
\begin{align*}
\lefteqn{\sum_{t=1}^T (z_{t,i}-m_{t,i})^2 - (z_{t,i} - m^*_i)^2} \\
 &\leq (m^*_i)^2 + \sum_{t=1}^T (z_{t,i}-m_{t,i})^2 - (z_{t,i} - m_{t+1,i})^2 \\
 &= (m^*_i)^2 + \sum_{t=1}^T 2z_{t,i}(m_{t+1,i}-m_{t,i}) + m_{t,i}^2 - m_{t+1,i}^2 \\
 &= (m^*_i)^2 + m_{1,i}^2 - m_{T+1,i}^2 + \sum_{t=1}^T \frac{2}{t+1}z_{t,i}(z_{t,i} - m_{t,i}) \\
 &\leq 1 + \sum_{t=1}^T \frac{z_{t,i}^2}{\gamma (t+1)^2} + \gamma(z_{t,i} - m_{t,i})^2 \\
 &\leq 1 + \frac{1}{\gamma} + \gamma \sum_{t=1}^T (z_{t,i} - m_{t,i})^2.
\end{align*}
Re-arranging yields
\begin{equation*}
\sum_{t=1}^T (z_{t,i} - m_{t,i})^2 \leq \frac{1}{1-\gamma}\left(\frac{1+\gamma}{\gamma} + \sum_{t=1}^T (z_{t,i} - m_i^*)^2\right),
\end{equation*}
which for $\gamma = \frac{1}{2}$ yields the desired result.
\end{proof}
Combining Proposition~\ref{prop:ftrl-aux} with Proposition~\ref{prop:aeg} yields 
the algorithm titled EG-Variance in Table~\ref{tab:algorithms}. This algorithm has the 
same updates as the algorithm given by \cite{hazan2010variation}, but uses the 
optimistic prediction $\exp(\theta_t - \eta m_t)$ rather than $\exp(\theta_t)$. 
As a straightforward consequence of Propositions~\ref{prop:aeg} and~\ref{prop:ftrl-aux}, we 
obtain the following regret bound, which improves the bound presented in 
\cite{hazan2010variation}:
\begin{corollary}
\label{cor:EG-variance}
For $0 < \eta \leq \frac{1}{4}$ and $\|z_t\|_{\infty} \leq 1$, the EG-Variance 
algorithm achieves a regret bound of
%\[ \Regret(w) \leq \frac{1 + (\log(n)-1)\|w\|_1 + \sum_{i=1}^n w_i\log(w_i)}{\eta} + \eta \left[ \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_i^*)^2 + 4\|w\|_1\log(T+1) \right]. \]
\begin{equation*}\Regret(w) \leq \frac{\log(n)}{\eta} + \eta \left[ 2\sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_i^*)^2 + 6 \right].\end{equation*} 
\end{corollary}

### 2014-01-27 lp/machinery.tex
proofread a first draft, now just need to add a conclusion

The regret bound for Algorithm~\ref{alg:adaptive-optimistic} is 
given in Theorem~\ref{thm:adaptive-optimistic}. A key observation is that 
the regret is in terms of $z_t - m_t$ rather than $z_t$ (so the closer 
$m_t$ is to $z_t$, the smaller the regret; this will be more apparent 
in various corollaries).
\begin{theorem}
\label{thm:adaptive-optimistic}
Algorithm~\ref{alg:adaptive-optimistic} achieves regret bounded by
\begin{dmath}
\label{eqn:regret-1}
\frac{1}{\eta}\left(\psi_1^*(\theta_1) + \psi_{T+1}(w) + \sum_{t=1}^{T} \psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t - \eta m_t) + \eta w_t^{\top}(z_t-m_t)\right).
\end{dmath}
\end{theorem}
\begin{proof}
First note that $\psi_t^*$ is convex and that 
$w_t = \nabla \psi_t^*(\theta_t - \eta m_t)$. It thus follows that 
$\psi_t^*(\theta_t) \geq \psi_t^*(\theta_t - \eta m_t) + \eta m_t^{\top}z_t$. 
Then, by definition of the Fenchel conjugate together with telescoping sums, we 
have, for any $w$, 
\begin{align*}
\lefteqn{\hskip -0.1in w^{\top}\theta_{T+1} - \psi_{T+1}(w)} \\
\hskip 0.1in  &\leq \psi_{T+1}^*(\theta_{T+1}) \\
  &= \psi_1^*(\theta_1) + \sum_{t=1}^{T} \psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t) \\
  &\leq \psi_1^*(\theta_1) + \sum_{t=1}^{T} \psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t-\eta m_t) - \eta w_t^{\top}m_t.
\end{align*}
Re-arranging and expanding $\theta_{T+1}$, we have
\begin{dmath}
-\eta w^{\top} \sum_{t=1}^T z_t \leq \psi_{T+1}(w) + \psi_1^*(\theta_1) + \sum_{t=1}^{T} \psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t-\eta m_t) - \eta w_t^{\top}m_t.
\end{dmath}
Adding $\eta \sum_{t=1}^T w_t^{\top}z_t$ to both sides, we obtain
\begin{dmath}
\eta \Regret(w) \leq \psi_{T+1}(w) + \psi_1^*(\theta_1) + \sum_{t=1}^T \psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t) + \eta w_t^{\top}(z_t-m_t),
\end{dmath}
which yields the desired result.
\end{proof}
As an immediate consequence, we have
\begin{corollary}
\label{cor:a-o1}
Suppose that $\psi_{t+1}$ is chosen so that $\psi_{t+1}^*(\theta_t - \eta z_t) \leq \psi_t^*(\theta_t - \eta m_t) - \eta w_t^{\top}(z_t - m_t)$. Then
\begin{equation}
\Regret(w) \leq \frac{\psi_1^*(\theta_1) + \psi_{T+1}(w)}{\eta}.
\end{equation}
\end{corollary}
\begin{proof}
Since $\theta_{t+1} = \theta_t - \eta z_t$, if the condition in the corollary 
holds then the sum in the regret bound (\ref{eqn:regret-1}) is bounded above 
term-wise by zero. 
Hence we obtain $\Regret(w) \leq \frac{1}{\eta}\left(\psi_1^*(\theta_1) + \psi_{T+1}(w)\right)$, as was to be shown.
\end{proof}
Finally, we consider a simplified version of Algorithm~\ref{alg:adaptive-optimistic}, 
which corresponds to choosing $\psi_t(w) = \psi(w) + \eta^2 w^{\top} \sum_{s=1}^{t-1} u_s$ 
for some fixed function $\psi$. In this case, we define 
$\psi_S^*(w) \eqdef \sup_{w \in S} w^{\top}x - \psi(w)$, which 
will allow us to more easily handle restrictions on the domain $S$.

@@

\paragraph{Application to exponentiated gradient.} We now show how to apply 
Corollary~\ref{cor:a-o2} to obtain an adaptive exponentiated gradient algorithm. 
We will state it here in the case of the simplex, since the bounds in this 
case are particularly easy to interpret. Note, however, that our analysis 
goes via Corollary~\ref{cor:a-o2}, and hence applies to arbitrary convex 
constraints.
\begin{proposition}[Adaptive Exponentiated Gradient]
\label{prop:aeg}
Consider the updates given by $\theta_{1,i} = 0$ and 
$\theta_{t+1,i} = \theta_{t,i} - \eta z_{t,i} - \eta^2 (z_{t,i} - m_{t,i})^2$, 
with prediction $w_{t,i} \propto \exp(\theta_{t,i} - \eta m_{t,i})$. Then, 
assuming $\|z_t\|_{\infty} \leq 1$, $\|m_t\|_{\infty} \leq 1$ and 
$0 < \eta \leq \frac{1}{4}$, we have
\begin{equation}
\Regret(w) \leq \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_{t,i})^2.
\end{equation}
\end{proposition}
\begin{proof}
We note that, for $S$ equal to the simplex $\Delta_n$ and 
$\psi(w) = \sum_{i=1}^n w_i\log(w_i)$, the $w_t$ is equal to 
$\nabla \psi_S^*(\theta_t - \eta m_t)$. It therefore suffices 
to check that the condition of Corollary~\ref{cor:a-o2} is satisfied with 
$u_{t,i} = (z_{t,i}-m_{t,i})^2$. 
Note that $\psi^*(\theta) = \sum_{i=1}^n \exp(\theta_i)$. We then have
\begin{dgroup*}
\begin{dmath*} \lefteqn{\psi^*(\theta_{t} - \eta z_t - \eta^2 u_t)} 
\end{dmath*}\begin{dmath*}  = \sum_{i=1}^n \exp(\theta_{t,i} - \eta z_{t,i} - \eta^2 (z_{t,i} - m_{t,i})^2) 
\end{dmath*}\begin{dmath*}  = \sum_{i=1}^n \exp(\theta_{t,i} - \eta m_{t,i})\exp(-\eta (z_{t,i} - m_{t,i}) - \eta^2 (z_{t,i} - m_{t,i})^2) 
\end{dmath*}\begin{dmath*}  \leq \sum_{i=1}^n \exp(\theta_{t,i} - \eta m_{t,i})(1 - \eta (z_{t,i} - m_{t,i})) 
\end{dmath*}\begin{dmath*}  = \sum_{i=1}^n \exp(\theta_{t,i} - \eta m_{t,i}) - \eta \sum_{i=1}^n \exp(\theta_{t,i} - \eta m_{t,i}) (z_{t,i} - m_{t,i}) 
\end{dmath*}\begin{dmath*}  = \psi^*(\theta_t - \eta m_t) - \eta \nabla \psi^*(\theta_t - \eta m_t)^{\top} (z_t - m_t).
\end{dmath*}
\end{dgroup*}
The one inequality we made use of was $\exp(-x-x^2) \leq 1-x$ for $|x| < \frac{1}{2}$.
This verifies the condition of Corollary~\ref{cor:a-o2}, yielding a regret bound 
of $\frac{\psi_S^*(0) + \psi(w)}{\eta} + \eta \sum_{i=1}^n w^{\top} u_t$. 
Finally, we note that $\psi_S^*(0) = \log(n)$, 
$\psi(w) = \sum_{i=1}^n w_i\log(w_i) \leq 0$, and 
$u_{t,i} = (z_{t,i} - m_{t,i})^2$, which completes the proof.
\end{proof}
Note that in the case that $m_{t,i} = 0$, the update is 
$\theta_{t+1,i} = \theta_{t,i} - \eta z_{t,i} - \eta^2 z_{t,i}^2$, whereas multiplicative 
updates by $1 - \eta z_{t,i}$ would correspond to 
$\theta_{t+1,i} = \theta_{t,i} + \log(1 - \eta z_{t,i})$. Since 
$-\eta z_{t,i} - \eta^2 z_{t,i}^2 \leq \log(1 - \eta z_{t,i})$ when 
$|\eta z_{t,i}| \leq \frac{1}{2}$, we can think of these updates as a second-order 
under-approximation to (\ref{eqn:mw2}), which however can more naturally incorporate 
the optimistic updates based on $m_t$.

>>>>

The regret bound for Algorithm~\ref{alg:adaptive-optimistic} is 
given in Theorem~\ref{thm:adaptive-optimistic}. A key observation is that 
the regret is in terms of $z_t - m_t$ rather than $z_t$ (so the closer 
$m_t$ is to $z_t$, the smaller the regret; this will be more apparent 
in various corollaries).
\begin{theorem}
\label{thm:adaptive-optimistic}
Algorithm~\ref{alg:adaptive-optimistic} achieves regret bounded by
\begin{dmath}
\label{eqn:regret-1}
\frac{1}{\eta}\left(\psi_1^*(\theta_1) + \psi_{T+1}(w) + \sum_{t=1}^{T} \left[\psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t - \eta m_t) + \eta w_t^{\top}(z_t-m_t)\right]\right).
\end{dmath}
\end{theorem}
\begin{proof}
First note that $\psi_t^*$ is convex and that 
$w_t = \nabla \psi_t^*(\theta_t - \eta m_t)$. It thus follows that 
$\psi_t^*(\theta_t) \geq \psi_t^*(\theta_t - \eta m_t) + \eta m_t^{\top}z_t$. 
Then, by definition of the Fenchel conjugate together with telescoping sums, we 
have, for any $w$, 
\begin{align*}
\lefteqn{\hskip -0.1in w^{\top}\theta_{T+1} - \psi_{T+1}(w)} \\
\hskip 0.1in  &\leq \psi_{T+1}^*(\theta_{T+1}) \\
  &= \psi_1^*(\theta_1) + \sum_{t=1}^{T} \psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t) \\
  &\leq \psi_1^*(\theta_1) + \sum_{t=1}^{T} \psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t-\eta m_t) - \eta w_t^{\top}m_t.
\end{align*}
Re-arranging and expanding $\theta_{T+1}$, we have
\begin{dmath}
-\eta w^{\top} \sum_{t=1}^T z_t \leq \psi_{T+1}(w) + \psi_1^*(\theta_1) + \sum_{t=1}^{T} [\psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t-\eta m_t) - \eta w_t^{\top}m_t].
\end{dmath}
Adding $\eta \sum_{t=1}^T w_t^{\top}z_t$ to both sides, we obtain
\begin{dmath}
\eta \Regret(w) \leq \psi_{T+1}(w) + \psi_1^*(\theta_1) + \sum_{t=1}^T [\psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t) + \eta w_t^{\top}(z_t-m_t)],
\end{dmath}
which yields the desired result.
\end{proof}
As an immediate consequence, we have
\begin{corollary}
\label{cor:a-o1}
Suppose that $\psi_{t+1}$ is chosen so that $\psi_{t+1}^*(\theta_t - \eta z_t) \leq \psi_t^*(\theta_t - \eta m_t) - \eta w_t^{\top}(z_t - m_t)$. Then
\begin{equation}
\Regret(w) \leq \frac{\psi_1^*(\theta_1) + \psi_{T+1}(w)}{\eta}.
\end{equation}
\end{corollary}
\begin{proof}
Since $\theta_{t+1} = \theta_t - \eta z_t$, if the condition in the corollary 
holds then the sum in the regret bound (\ref{eqn:regret-1}) is bounded above 
term-wise by zero. 
Hence we obtain $\Regret(w) \leq \frac{1}{\eta}\left(\psi_1^*(\theta_1) + \psi_{T+1}(w)\right)$, as was to be shown.
\end{proof}
Finally, we consider a simplified version of Algorithm~\ref{alg:adaptive-optimistic}, 
which corresponds to choosing $\psi_t(w) = \psi(w) + \eta^2 w^{\top} \sum_{s=1}^{t-1} u_s$ 
for some fixed function $\psi$. In this case, we define 
$\psi_S^*(w) \eqdef \sup_{w \in S} w^{\top}x - \psi(w)$, which 
will allow us to more easily handle restrictions on the domain $S$.

@@

\paragraph{Application to exponentiated gradient.} We now show how to apply 
Corollary~\ref{cor:a-o2} to obtain an adaptive exponentiated gradient algorithm. 
We will state it here in the case of the simplex, since the bounds in this 
case are particularly easy to interpret. Note, however, that our analysis 
goes via Corollary~\ref{cor:a-o2}, and hence applies to arbitrary convex 
constraints.
\begin{proposition}[Adaptive Exponentiated Gradient]
\label{prop:aeg}
Consider the updates given by $\theta_{1,i} = 0$ and 
$\theta_{t+1,i} = \theta_{t,i} - \eta z_{t,i} - \eta^2 (z_{t,i} - m_{t,i})^2$, 
with prediction $w_{t,i} \propto \exp(\theta_{t,i} - \eta m_{t,i})$. Then, 
assuming $\|z_t\|_{\infty} \leq 1$, $\|m_t\|_{\infty} \leq 1$ and 
$0 < \eta \leq \frac{1}{4}$, we have
\begin{equation}
\Regret(w) \leq \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_{t,i})^2.
\end{equation}
\end{proposition}
\begin{proof}
We note that, for $S$ equal to the simplex $\Delta_n$ and 
$\psi(w) = \sum_{i=1}^n w_i\log(w_i)$, $w_t$ is equal to 
$\nabla \psi_S^*(\theta_t - \eta m_t)$. It therefore suffices 
to check that the condition of Corollary~\ref{cor:a-o2} is satisfied with 
$u_{t,i} = (z_{t,i}-m_{t,i})^2$. 
Note that $\psi^*(\theta) = \sum_{i=1}^n \exp(\theta_i)$. We then have
\begin{dgroup*}
\begin{dmath*} \lefteqn{\psi^*(\theta_{t} - \eta z_t - \eta^2 u_t)} 
\end{dmath*}\begin{dmath*}  = \sum_{i=1}^n \exp(\theta_{t,i} - \eta z_{t,i} - \eta^2 (z_{t,i} - m_{t,i})^2) 
\end{dmath*}\begin{dmath*}  = \sum_{i=1}^n \exp(\theta_{t,i} - \eta m_{t,i})\exp(-\eta (z_{t,i} - m_{t,i}) - \eta^2 (z_{t,i} - m_{t,i})^2) 
\end{dmath*}\begin{dmath*}  \leq \sum_{i=1}^n \exp(\theta_{t,i} - \eta m_{t,i})(1 - \eta (z_{t,i} - m_{t,i})) 
\end{dmath*}\begin{dmath*}  = \sum_{i=1}^n \exp(\theta_{t,i} - \eta m_{t,i}) - \eta \sum_{i=1}^n \exp(\theta_{t,i} - \eta m_{t,i}) (z_{t,i} - m_{t,i}) 
\end{dmath*}\begin{dmath*}  = \psi^*(\theta_t - \eta m_t) - \eta \nabla \psi^*(\theta_t - \eta m_t)^{\top} (z_t - m_t).
\end{dmath*}
\end{dgroup*}
The one inequality we made use of was $\exp(-x-x^2) \leq 1-x$ for $|x| < \frac{1}{2}$.
This verifies the condition of Corollary~\ref{cor:a-o2}, yielding a regret bound 
of $\frac{\psi_S^*(0) + \psi(w)}{\eta} + \eta \sum_{i=1}^n w^{\top} u_t$. 
Finally, we note that $\psi_S^*(0) = \log(n)$, 
$\psi(w) = \sum_{i=1}^n w_i\log(w_i) \leq 0$, and 
$u_{t,i} = (z_{t,i} - m_{t,i})^2$, which completes the proof.
\end{proof}
Note that in the case that $m_{t,i} = 0$, the update is 
$\theta_{t+1,i} = \theta_{t,i} - \eta z_{t,i} - \eta^2 z_{t,i}^2$, whereas multiplicative 
updates by $1 - \eta z_{t,i}$ would correspond to 
$\theta_{t+1,i} = \theta_{t,i} + \log(1 - \eta z_{t,i})$. Since 
$-\eta z_{t,i} - \eta^2 z_{t,i}^2 \leq \log(1 - \eta z_{t,i})$ when 
$|\eta z_{t,i}| \leq \frac{1}{2}$, we can think of these updates as a second-order 
under-approximation to (\ref{eqn:mw2}), which however can more naturally incorporate 
the optimistic updates based on $m_t$.
